A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Updates from Overleaf

\section*{DISCUSSION}
There are two major contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analysis widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
if we like this].
We hope that the ARG provided here will similarly
enable a range recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements fully integrating
recombination into pathogen genomics.

The second major contribution is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
By capturing recombination in real time at a global scale, our approach extends the established utility of comprehensive genealogies~\cite{Hinrichs2024,McBroome2024}, offers clear insight into recombination dynamics, and supports timely investigation of .
The novelty of sc2ts and the inferred ARG motivates our  focus on methodological details. 
Through these detailed analyses we validate the ARG itself,
ensuring that it is a trustworthy basis for future research,
and deepen our understanding of the inference method, and 
how it may be adapted and improved (see Limitations of the Study).

Study of the sc2ts ARG provides insights in the  contributions of this work are provided by our 
insights into the evolution of SARS-CoV-2 obtained by studying
the sc2ts ARG. Firstly, we have clarified the relationships among
the recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Secondly, we have used the ARG to study breakpoint intervals 
for recombination events, adding a novel temporal dimension
and demonstrating a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage heterozygosity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we have provided a stringent lower
bound on the amount of recombination that occured during
the pandemic phase of SARS-CoV-2.
Together, these  underscore the utility of the ARG itself, and suggest
several important directions for future research.

@@

% SHZ: WIP
We close by emphasizing several current limitations
and potential directions to improve sc2ts,
which we hope will also inspire future research
to develop ARG inference methods for pathogens beyond SARS-CoV-2.

>>>>

\section*{DISCUSSION}
There are two major contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analysis widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
if we like this].
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements which fully integrate
recombination into pathogen genomics.

The second major contribution is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
By capturing recombination in real time at a global scale, our approach extends the established utility of comprehensive genealogies~\cite{Hinrichs2024,McBroome2024}, providing clear insight and enabling timely investigation of recombination-aware pathogen dynamics.
The novelty of the inferred ARG motivates our detailed analyses of the methodology behind sc2ts, which we use to validate the ARG itself. This ensure that the ARG is a trustworthy basis for future research, deepens our understanding of the inference method, and suggests avenues for improvement (see Limitations of the Study).

Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, and this work therefore makes several additional contributions to the field. First, we clarify the relationships among
recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Second, we use the ARG to study breakpoint intervals 
for recombination events, add a novel temporal dimension
and show a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage heterozygosity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we  provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
Together, these  underscore the utility of the ARG itself, and suggest
several important directions for future research.

@@

% SHZ: WIP
We close by emphasizing several current limitations of sc2ts
and suggest potential directions for improvement
which we hope will inspire future development of ARG inference
methods for pathogens beyong SARS-CoV2.
